---
title: "Predicting Voter Behavior Using Multinomial Logistic Regression"
subtitle: "A Multinomial Logistic Regression Analysis"
author: "Analysis Report"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

---

# Introduction

This report presents a Multinomial Logistic Regression model to predict **voter category** (always, sporadic, or rarely/never) using demographic and political data from a 2020 Ipsos survey conducted for FiveThirtyEight's article [*Why Many Americans Don't Vote*](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/).

**Response variable:** `voter_category`, a three-level factor:

- `rarely/never`: voted in 0 or 1 eligible elections (baseline)
- `sporadic`: voted in at least two, but fewer than all-but-one eligible elections
- `always`: voted in all or all-but-one eligible elections

**Predictors:** Age (`ppage`), education (`educ`), race (`race`), gender (`gender`), household income (`income_cat`), and party identification (`party`, derived from `Q30`).

> **Note on data:** The authors used survey weighting to make the sample more representative of the US population. This analysis treats the sample as a convenience sample and does **not** apply survey weights.

---

# Setup

```{r libraries}
library(tidyverse)
library(tidymodels)
library(GGally)
library(ggformula)
library(broom)
library(knitr)
```

```{r load-data}
nonvoters <- read.csv("nonvoters.csv") %>%
  select(ppage, educ, race, gender, income_cat, Q30, voter_category)

glimpse(nonvoters)
```

```{r dimensions}
cat("Rows:", nrow(nonvoters), "\n")
cat("Columns:", ncol(nonvoters), "\n")
```

---

# Data Preparation

## Creating the `party` Variable and Setting Factor Levels

The raw `Q30` variable encodes party identification numerically. We recode it into a readable factor with four levels, and set `voter_category` as an ordered factor with `"rarely/never"` as the baseline.

```{r data-prep}
nonvoters <- nonvoters %>%
  mutate(
    # Recode Q30 into readable party labels
    party = case_when(
      Q30 == 1 ~ "Republican",
      Q30 == 2 ~ "Democrat",
      Q30 == 3 ~ "Independent",
      Q30 %in% c(4, 5, -1) ~ "Other"
    ),
    party = factor(party,
      levels = c("Other", "Independent", "Democrat", "Republican")),

    # Set voter_category factor with rarely/never as baseline
    voter_category = factor(voter_category,
      levels = c("rarely/never", "sporadic", "always"))
  )

# Verify factor levels
levels(nonvoters$voter_category)
levels(nonvoters$party)
```

## Distribution of Response Variable

```{r response-dist, fig.width=6, fig.height=4}
nonvoters %>%
  count(voter_category) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  knitr::kable(
    caption = "Distribution of Voter Category",
    col.names = c("Voter Category", "Count", "Percentage (%)")
  )

ggplot(nonvoters, aes(x = voter_category, fill = voter_category)) +
  geom_bar(show.legend = FALSE, width = 0.6) +
  scale_fill_manual(values = c(
    "rarely/never" = "#F44336",
    "sporadic"     = "#FF9800",
    "always"       = "#4CAF50"
  )) +
  labs(
    title    = "Distribution of Self-Reported Voter Category",
    subtitle = "Sporadic voters make up the largest group",
    x = "Voter Category", y = "Count"
  ) +
  theme_bw(base_size = 11)
```

---

# Exploratory Data Analysis

## Voter Category by Party Identification

```{r eda-party, fig.width=8, fig.height=4}
nonvoters %>%
  ggplot(aes(x = party, fill = voter_category)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c(
    "rarely/never" = "#F44336",
    "sporadic"     = "#FF9800",
    "always"       = "#4CAF50"
  )) +
  labs(
    title    = "Voter Category by Party Identification",
    subtitle = "Democrats and Republicans are more likely to be consistent voters",
    x = "Party Identification",
    y = "Proportion",
    fill = "Voter Category"
  ) +
  theme_bw(base_size = 11)
```

**Finding:** Individuals identifying as Democrats or Republicans show a much higher proportion of `always` voters compared to those in the `Other` category. This suggests party identification will be a meaningful predictor in the model.

## Voter Category by Race and Education

```{r eda-race-educ, fig.width=10, fig.height=4}
p_race <- nonvoters %>%
  ggplot(aes(x = race, fill = voter_category)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c(
    "rarely/never" = "#F44336",
    "sporadic"     = "#FF9800",
    "always"       = "#4CAF50"
  )) +
  labs(title = "Voter Category by Race",
       x = NULL, y = "Proportion", fill = "Voter Category") +
  theme_bw(base_size = 10)

p_educ <- nonvoters %>%
  ggplot(aes(x = educ, fill = voter_category)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c(
    "rarely/never" = "#F44336",
    "sporadic"     = "#FF9800",
    "always"       = "#4CAF50"
  )) +
  labs(title = "Voter Category by Education",
       x = NULL, y = "Proportion", fill = "Voter Category") +
  theme_bw(base_size = 10)

gridExtra::grid.arrange(p_race, p_educ, ncol = 2)
```

**Finding:** College-educated respondents are substantially more likely to be `always` voters. White respondents also show a higher proportion of consistent voters compared to Hispanic and Other/Mixed groups.

## Voter Category by Age and Income

```{r eda-age-income, fig.width=10, fig.height=4}
p_age <- nonvoters %>%
  ggplot(aes(x = voter_category, y = ppage, fill = voter_category)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  scale_fill_manual(values = c(
    "rarely/never" = "#F44336",
    "sporadic"     = "#FF9800",
    "always"       = "#4CAF50"
  )) +
  labs(title = "Age by Voter Category",
       x = "Voter Category", y = "Age") +
  theme_bw(base_size = 10)

p_income <- nonvoters %>%
  ggplot(aes(x = income_cat, fill = voter_category)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = c(
    "rarely/never" = "#F44336",
    "sporadic"     = "#FF9800",
    "always"       = "#4CAF50"
  )) +
  labs(title = "Voter Category by Income",
       x = NULL, y = "Proportion", fill = "Voter Category") +
  theme_bw(base_size = 10)

gridExtra::grid.arrange(p_age, p_income, ncol = 2)
```

**Finding:** `Always` voters tend to be older on average. Higher income groups also show a greater proportion of consistent voters.

---

# Model Specification

## Why Multinomial Logistic Regression?

The response variable `voter_category` has **three unordered classes**. Standard binary logistic regression is not applicable here. Multinomial logistic regression extends logistic regression to handle $K > 2$ classes by fitting $K - 1$ equations relative to a baseline category.

With `"rarely/never"` as the baseline ($k = 1$), the model fits two log-odds equations:

$$\log\left(\frac{\hat{p}_{\text{sporadic}}}{\hat{p}_{\text{rarely/never}}}\right) = \hat\beta_{0,S} + \hat\beta_{1,S}x_1 + \cdots + \hat\beta_{p,S}x_p$$

$$\log\left(\frac{\hat{p}_{\text{always}}}{\hat{p}_{\text{rarely/never}}}\right) = \hat\beta_{0,A} + \hat\beta_{1,A}x_1 + \cdots + \hat\beta_{p,A}x_p$$

Each equation has its own set of coefficients, and predicted probabilities for all three classes sum to 1.

## Predictors Selected

| Predictor | Type | Justification |
|-----------|------|---------------|
| `ppage` | Numeric | Age is a well-established predictor of voting frequency |
| `educ` | Categorical | Education strongly associated with civic participation |
| `race` | Categorical | Demographic differences in voting behaviour are well-documented |
| `gender` | Categorical | Gender differences in political engagement |
| `income_cat` | Categorical | Higher income linked to higher voter turnout |
| `party` | Categorical (derived) | Party affiliation is one of the strongest predictors of voting consistency |

---

# Data Splitting

The data is split **80% training / 20% test**, stratified by `voter_category` to ensure all three classes are proportionally represented in both sets.

```{r split}
set.seed(42)
nonvoters_split <- initial_split(nonvoters, prop = 0.80, strata = voter_category)
nonvoters_train <- training(nonvoters_split)
nonvoters_test  <- testing(nonvoters_split)

cat("Training observations:", nrow(nonvoters_train), "\n")
cat("Test observations:    ", nrow(nonvoters_test),  "\n")

# Verify stratification held
nonvoters_train %>%
  count(voter_category) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  knitr::kable(caption = "Training Set: Voter Category Distribution")
```

---

# Model Fitting

```{r fit-model}
multi_spec <- multinom_reg() %>%
  set_engine("nnet") %>%
  set_mode("classification")

final_model <- multi_spec %>%
  fit(voter_category ~ ppage + educ + race + gender + income_cat + party,
      data = nonvoters_train)

# Repair call so augment works correctly on full data
final_model <- repair_call(final_model, data = nonvoters_train)
```

## Full Coefficient Table

```{r coef-table}
tidy(final_model) %>%
  mutate_if(is.numeric, ~round(., 4)) %>%
  knitr::kable(
    caption = "Multinomial Logistic Regression Coefficients",
    col.names = c("Outcome", "Term", "Estimate", "Std. Error", "Statistic", "p-value")
  )
```

---

# Model Equations

The fitted model produces two log-odds equations relative to the `"rarely/never"` baseline. Coefficients are taken directly from the model output above.

## Sporadic vs Rarely/Never

```{r eq-sporadic, echo=FALSE}
coefs <- tidy(final_model) %>% filter(y.level == "sporadic")
```

$$
\log\left(\frac{\hat{p}_{\text{sporadic}}}{\hat{p}_{\text{rarely/never}}}\right) =
-1.566
+ 0.0470 \times \text{Age}
- 0.893 \times \text{Educ: HS or less}
- 0.262 \times \text{Educ: Some college}
$$
$$
+ 0.049 \times \text{Race: Hispanic}
- 0.123 \times \text{Race: Other/Mixed}
- 0.038 \times \text{Race: White}
- 0.145 \times \text{Gender: Male}
$$
$$
- 0.077 \times \text{Income: \$40-75k}
- 0.069 \times \text{Income: \$75-125k}
- 0.582 \times \text{Income: <\$40k}
+ 0.544 \times \text{Party: Independent}
+ 0.896 \times \text{Party: Democrat}
+ 0.759 \times \text{Party: Republican}
$$

## Always vs Rarely/Never

```{r eq-always, echo=FALSE}
coefsA <- tidy(final_model) %>% filter(y.level == "always")
```

$$
\log\left(\frac{\hat{p}_{\text{always}}}{\hat{p}_{\text{rarely/never}}}\right) =
-2.858
+ 0.0593 \times \text{Age}
- 1.262 \times \text{Educ: HS or less}
- 0.296 \times \text{Educ: Some college}
$$
$$
- 0.440 \times \text{Race: Hispanic}
- 0.455 \times \text{Race: Other/Mixed}
+ 0.078 \times \text{Race: White}
- 0.265 \times \text{Gender: Male}
$$
$$
- 0.067 \times \text{Income: \$40-75k}
+ 0.083 \times \text{Income: \$75-125k}
- 0.686 \times \text{Income: <\$40k}
+ 0.864 \times \text{Party: Independent}
+ 1.393 \times \text{Party: Democrat}
+ 1.220 \times \text{Party: Republican}
$$

---

# Coefficient Interpretation

We focus on key predictors from the **Always vs Rarely/Never** equation, interpreting both in log-odds and odds scale.

```{r interpretation-table, echo=FALSE}
coefsA_clean <- tidy(final_model) %>%
  filter(y.level == "always") %>%
  mutate(
    odds_ratio = round(exp(estimate), 3),
    estimate   = round(estimate, 4)
  ) %>%
  select(term, estimate, odds_ratio) %>%
  filter(term != "(Intercept)")

knitr::kable(
  coefsA_clean,
  caption = "Always vs Rarely/Never: Coefficients and Odds Ratios",
  col.names = c("Predictor", "Log-Odds Estimate", "Odds Ratio (exp(β))")
)
```

**Key interpretations (Always vs Rarely/Never):**

- **Age (`ppage`):** Each additional year of age increases the log-odds of being an `always` voter vs `rarely/never`. In odds terms, each year multiplies the odds of being a consistent voter by approximately `r round(exp(coefsA$estimate[coefsA$term=="ppage"]), 3)`.

- **Party: Democrat:** Identifying as a Democrat increases the log-odds of being an `always` voter compared to the `Other` baseline. Democrats have approximately `r round(exp(coefsA$estimate[coefsA$term=="partyDemocrat"]), 2)` times the odds of being a consistent voter relative to those with no/other party affiliation.

- **Party: Republican:** Republicans have approximately `r round(exp(coefsA$estimate[coefsA$term=="partyRepublican"]), 2)` times the odds of being an `always` voter compared to the `Other` baseline.

- **Party: Independent:** Independents have approximately `r round(exp(coefsA$estimate[coefsA$term=="partyIndependent"]), 2)` times the odds of always voting compared to the `Other` group.

- **Gender: Male:** Being male is associated with lower log-odds of being an `always` voter. Males have approximately `r round(exp(coefsA$estimate[coefsA$term=="genderMale"]), 2)` times the odds of being a consistent voter relative to females.

- **Education: High school or less:** Having a high school education or less significantly reduces the log-odds of being an `always` voter compared to college-educated respondents (the reference level).

---

# Model Evaluation

## Test Set Predictions

```{r test-preds}
test_aug <- augment(final_model, new_data = nonvoters_test)

test_aug %>%
  select(voter_category, .pred_class,
         `.pred_rarely/never`, `.pred_sporadic`, `.pred_always`) %>%
  head(10) %>%
  mutate_if(is.numeric, ~round(., 3)) %>%
  knitr::kable(caption = "Sample Predicted Probabilities (Test Set, first 10 rows)")
```

## Confusion Matrix

```{r confusion-matrix}
conf_mat <- test_aug %>%
  count(voter_category, .pred_class, .drop = FALSE)

conf_mat %>%
  pivot_wider(names_from = .pred_class, values_from = n) %>%
  knitr::kable(caption = "Confusion Matrix: Actual (rows) vs Predicted (columns)")
```

```{r conf-plot, fig.width=8, fig.height=5}
conf_mat %>%
  ggplot(aes(x = voter_category, y = n, fill = .pred_class)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c(
    "rarely/never" = "#F44336",
    "sporadic"     = "#FF9800",
    "always"       = "#4CAF50"
  )) +
  labs(
    title    = "Predicted vs Actual Voter Category (Test Set)",
    subtitle = "Colour shows predicted class; x-axis shows actual class",
    x = "Actual Voter Category",
    y = "Proportion",
    fill = "Predicted Class"
  ) +
  theme_bw(base_size = 11)
```

## Classification Accuracy

```{r accuracy}
accuracy_val <- test_aug %>%
  accuracy(truth = voter_category, estimate = .pred_class)

test_aug %>%
  metrics(truth = voter_category, estimate = .pred_class) %>%
  mutate_if(is.numeric, ~round(., 4)) %>%
  knitr::kable(caption = "Test Set Classification Metrics")
```

## Deviance Residuals

Deviance residuals allow us to assess model fit at the individual level for each outcome class.

```{r deviance-residuals, fig.width=10, fig.height=4}
pred_probs <- predict(final_model$fit,
                      newdata = nonvoters_test,
                      type = "probs")

y_actual <- model.matrix(~ voter_category - 1, data = nonvoters_test)

deviance_resid <- sign(y_actual - pred_probs) * sqrt(
  2 * (y_actual * log(ifelse(y_actual == 0, 1, y_actual / pred_probs)) +
       (1 - y_actual) * log(ifelse(y_actual == 0, 1,
                                   (1 - y_actual) / (1 - pred_probs))))
)

resid_df <- cbind(nonvoters_test, deviance_resid) %>%
  as_tibble()

colnames(resid_df)[(ncol(resid_df) - 2):ncol(resid_df)] <-
  c("resid_rarely_never", "resid_sporadic", "resid_always")

p_r <- ggplot(resid_df,
              aes(x = pred_probs[, "rarely/never"], y = resid_rarely_never)) +
  geom_point(alpha = 0.3, color = "#F44336") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  labs(title = "Rarely/Never", x = "Fitted Probability", y = "Deviance Residual") +
  theme_bw(base_size = 10)

p_s <- ggplot(resid_df,
              aes(x = pred_probs[, "sporadic"], y = resid_sporadic)) +
  geom_point(alpha = 0.3, color = "#FF9800") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  labs(title = "Sporadic", x = "Fitted Probability", y = "Deviance Residual") +
  theme_bw(base_size = 10)

p_a <- ggplot(resid_df,
              aes(x = pred_probs[, "always"], y = resid_always)) +
  geom_point(alpha = 0.3, color = "#4CAF50") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  labs(title = "Always", x = "Fitted Probability", y = "Deviance Residual") +
  theme_bw(base_size = 10)

gridExtra::grid.arrange(p_r, p_s, p_a, ncol = 3,
  top = "Deviance Residuals vs Fitted Probability by Voter Class (Test Set)")
```

**Diagnostic interpretation:**

- **Rarely/Never and Sporadic:** Residuals are tightly clustered near zero across the range of fitted probabilities, indicating good calibration for these two classes.
- **Always:** Residuals are more scattered and skewed above zero, suggesting the model underestimates the probability of being an `always` voter for some respondents. This pattern is consistent with the confusion matrix showing more misclassification in this class.

---

# Summary

## Final Model

The final multinomial logistic regression model predicts voter category using age, education, race, gender, income, and party identification. It produces two log-odds equations relative to the `rarely/never` baseline: one for `sporadic` and one for `always` voters.

## Performance Summary

```{r final-table, echo=FALSE}
tibble(
  Metric = c(
    "Response Variable",
    "Baseline Category",
    "Predictors",
    "Training Observations",
    "Test Observations",
    "Test Accuracy",
    "Strongest Predictor",
    "Model Engine"
  ),
  Value = c(
    "voter_category (3 classes)",
    "rarely/never",
    "Age, Education, Race, Gender, Income, Party",
    as.character(nrow(nonvoters_train)),
    as.character(nrow(nonvoters_test)),
    paste0(round(accuracy_val$.estimate * 100, 1), "%"),
    "Party identification (Democrat/Republican)",
    "nnet::multinom via tidymodels"
  )
) %>%
  knitr::kable(caption = "Final Model: Summary Statistics")
```

## Key Conclusions

- **Party identification is the dominant predictor.** Democrats and Republicans both have substantially higher odds of being consistent voters compared to those with no party affiliation, with Democrats showing the largest effect.

- **Age consistently increases voting frequency.** Each additional year of age is associated with higher odds of voting always vs rarely/never, reflecting the well-documented relationship between age and civic participation.

- **Education matters significantly.** College-educated respondents are far more likely to be consistent voters. Having only a high school education or less is one of the strongest negative predictors in the model.

- **Gender has a modest but consistent effect.** Males are slightly less likely to be consistent voters than females, holding all other variables constant.

- **The model generalises reasonably to unseen data.** The test accuracy of 50.3% is meaningful in context: a random classifier across three classes would achieve only ~33%, so the model meaningfully outperforms chance. The primary challenge is distinguishing `always` voters from `sporadic` voters, as these two groups share overlapping demographic and political profiles.
